```{r}
library(tidyverse)
library(readxl)
library(ggthemes)
library(RColorBrewer)
library(sf)
library(scales) # for percent function 

# Election data 
election <- read_excel("Data/kdl election data 2023.xlsx")

spatial <- st_read("Data/tl_2022_26_cousub/tl_2022_26_cousub.shp")

USA <- st_read("Data/tl_2022_us_county/tl_2022_us_county.shp")

#Reading in the Michigan county data 
michigan <- st_read("Data/tl_2018_us_county/tl_2018_us_county.shp") %>% 
      filter(STATEFP == 26)

#Just Kent county data 
kent <- filter(michigan, NAME == "Kent")
```
# Data Processing 

## string processing 
```{r}
# Gather township, city, and charter township info
election$govt <- character(length(election$Precinct))  # Initialize a new column

for (i in seq_along(election$Precinct)) {
  split_result <- str_split(election$Precinct[i], " ")
  
  # Adding charter township exception
  if (split_result[[1]][3] %in% c("Township", "Charter")) {
    election$govt[i] <- "Charter Township"
    
  # Handling exception for EGR  
  } else if (split_result[[1]][4] == "City") {
      election$govt[i] <- "City"
    
  # The rest     
  } else {
    election$govt[i] <- split_result[[1]][2]
  }
}

# this regex pattern will match any occurrence of the specified words ("Ward", "Precinct" or "Precincts", "Charter", "Township", "City"), the HTML entity for ampersand ("&"), and any digit (0-9) in a given text. The (s)? allows for an optional 's' in the "Precinct" part, making it flexible for both singular and plural forms.

pattern1 <- "Ward|Precinct(s)?|Charter|Township|City|&amp;|\\d"

pattern2 <- "Ward|Precinct(s)?|&amp;|\\d"

election$city_name <- str_trim(str_remove_all(election$Precinct, pattern = pattern1))

election$full_cn <- str_trim(str_remove_all(election$Precinct, pattern = pattern2))

rm(split_result)
```


## Condolsidating each precinct based on township/city + calculating new variables 
```{r}
election_result <- election %>% 
  group_by(full_cn) %>%
  summarise( 
    govt = first(govt), #the first function, which will retain the first non-missing value for each group (Precinct)
    total_registered_voters = sum(total_registered_voters),
    ballots_cast = sum(ballots_cast),
    turnout = round(sum(ballots_cast)/sum(total_registered_voters), 3), # I take 3 decimal places out to help with rounding errors later on
    yes = sum(yes),
    no = sum(no),
    
    #New additions 
    margin_raw = sum(yes) - sum(no),
    percent_yes = round(sum(yes)/sum(ballots_cast), 2),
    percent_no = round(sum(no)/sum(ballots_cast), 2),
    margin_percent = sum(percent_yes) - sum(percent_no))
```


# EDA
```{r}
vars <- list("percent_yes", "margin_percent", "turnout") 


for (col_name in vars) {
  cat("\nSummary Statistics for '", col_name, "':\n")
  print(summary(election_result[[col_name]]))
}

rm(vars)
```



```{r}
ggplot(election_result, aes(x = reorder(full_cn, total_registered_voters), y = total_registered_voters, fill = total_registered_voters)) +
  
  geom_bar(stat = "identity", color = "black") + 
  
  scale_y_continuous(
  breaks = seq(0, 60000, by = 10000),
  limits = c(0, 60000)) +

  scale_fill_gradient(low = "royalblue4", high = "springgreen3") +  # Color gradient from blue to green +

  labs(title = "Total Registered Voters by Locale",
       x = NULL,
       y = "Total Registered Voters",
       fill = NULL) +

  theme_clean() +
  
  theme(panel.grid.minor.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.y = element_blank(),
        legend.position = "none") +
          
  coord_flip() 


#display.brewer.all()
```


# Graphing voter turnout as proportional to total registered voters 
```{r}
ggplot(election_result, aes(x = reorder(full_cn, total_registered_voters))) +
  
  geom_bar(aes(y = total_registered_voters), stat = "identity", color = "black", fill = "#E31A1C") + 
  geom_bar(aes(y = ballots_cast), stat = "identity", color = "black", fill = "#33A02C") +
  
  geom_text(aes(y = ballots_cast, label = percent(turnout)), hjust = -0.2, size = 2.4, color = "black") + # Need to add the y aesthetics so R knows where to plot the label + used percent function from scales library to turn each numeric value into a percent 
  
  annotate("text", x = 16, y = 50000,
         label = paste0("Green = Ballots Cast\nRed = Total Registered Voters \n Voter Turnout Rate"),
         size = 3, color = "black", family = "Arial", fontface = "bold") + 

  scale_y_continuous(breaks = seq(0, 60000, by = 10000),
                    limits = c(0, 60000)) +
  
  #scale_fill_gradient(low = "royalblue4", high = "springgreen3") +  # Color gradient from blue to green +

  labs(title = "Ballots Cast vs Total Registered Votes",
       x = NULL,
       y = "Number of Voters",
       fill = NULL) +

  theme_clean() +
  
  theme(panel.grid.minor.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.y = element_blank()) +
          
  coord_flip() 

#ggsave(one, file = "Graphs/c123.png")


# Finding colorblind friendly colors + pallets

#display.brewer.all(colorblindFriendly=TRUE)
#display.brewer.pal(12, "Set2")
#brewer.pal(12, "Set2")
```



## Yes vote 
```{r}
avg_yes = sum(election_result$yes) / sum(election_result$ballots_cast)


ggplot(election_result, aes(x = reorder(full_cn, percent_yes), y = percent_yes, fill = percent_yes)) +
  geom_bar(stat = "identity", color = "black") +
  
  geom_hline(yintercept = avg_yes, color = "blue") +
  
  scale_fill_gradient(low = "#FC8D62", high = "#66C2A5") +  # Color gradient from blue to green
  
  scale_y_continuous(
    breaks = seq(0, 1, by = .2),
    limits = c(0, 1),
    labels = scales::percent_format(scale = 100)) +  # adding a label to the x variable to show a percentage
  
  labs(
    x = NULL, 
    y = "Yes Vote", 
    title = "Election Results for KDL's 2023 Millage by Area",
    subtitle = NULL) +
  
  theme_clean() +
  
  theme(panel.grid.minor.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.y = element_blank(),
        legend.position = "none") +
  
 coord_flip()
```


### Lollipop plot 
```{r}
ggplot(election_result, aes(x = reorder(full_cn, percent_yes), y = percent_yes)) +
  geom_point( color="blue", size=4, alpha=0.6) + # for reorder function to work geom_point needs to precede geom_segment
  geom_segment(aes(x=full_cn, xend=full_cn, y=0, yend=percent_yes), color="skyblue", size = 1) +
  
  scale_y_continuous(
  breaks = seq(0, 1, by = .2),
  limits = c(0, 1),
  labels = scales::percent_format(scale = 100)) +
  
  labs(
    x = NULL, 
    y = "Yes Vote", 
    title = "Election Results for KDL's 2023 Millage by Area",
    subtitle = NULL) +
  
  theme_clean() +
  
  coord_flip() +
  theme(panel.grid.minor.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
    panel.grid.major.y = element_blank())

#ggsave(one, file = "Graphs/c12433.png")

```




## Voter turnout rates 
```{r}
#looking at voter turnout rates 

#calculating average voter turnout across Kent county
avg_turnout = sum(election_result$ballots_cast) / sum(election_result$total_registered_voters)

ggplot(election_result, aes(x = reorder(full_cn, turnout ), y = turnout, fill = turnout)) +
  geom_bar(stat = "identity", color = "black") +
  
  geom_hline(yintercept = avg_turnout, color = "blue") +
  
  scale_fill_gradient(low = "#FC8D62", high = "#66C2A5") +  # Color gradient from blue to green
  
  scale_y_continuous(
    breaks = seq(0, 1, by = .2),
    limits = c(0, 1),
    labels = scales::percent_format(scale = 100)) +
  
  labs(
    x = NULL, 
    y = "Turnout Rate", 
    title = "Voter Turnout for 2023 November Election",
    subtitle = NULL) +
  
  theme_clean() +
  
  theme(panel.grid.minor.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
        panel.grid.major.y = element_blank(),
        legend.position = "none") +
  
 coord_flip()
```


# Mapping Analysis

## Merging voting data with spatial data  
```{r}
#Changing the case of observations in the spatial data to match election_result 
spatial$NAMELSAD <- str_to_title(spatial$NAMELSAD)

#Renaming the country name column to match column in election_result
spatial <- spatial %>% rename(full_cn = NAMELSAD) %>% filter(COUNTYFP == "081")

#Little discrepency between Caledonia CT and just township
spatial[5, "full_cn"] <- "Caledonia Charter Township"

#Merging the datasets based on the full township/county/city name to avoid duplicates with same names  
merged_data <- merge(spatial, election_result, by = "full_cn") 

#merged_data <- merged_data %>% filter(COUNTYFP == "081")
```


```{r}
#Slightly editing the merged_data to display more relevant data

merged_data[c(12, 16, 17), "NAME"] <- c("GR CT", "Lowell CT", "Lowell City")
```


## Kent county map data for Yes Vote (with missing data)
```{r}
ggplot() +
  geom_sf(data = kent, fill = "lightgrey", alpha = 0.5) +
  geom_sf(data = merged_data, aes(fill = percent_yes)) +
  
  #Adding labels to each county subdivision 
  geom_sf_text(data = merged_data, aes(label = NAME), color = "black", size = 2.2) +
  
  scale_fill_distiller(palette = "Spectral", trans = "reverse") +
  
  labs(title = "Kent County Voting Data for KDL Millage Renewel",
       subtitle = "Grey Areas not within KDL",
       fill = "Percent of Yes Votes",
       caption = "Source: Kent Election Reporting") +
  
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), #hjust = 0.5 centers the title
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 8, colour = "black"),
        plot.caption = element_text(hjust = 1.5, vjust = 5)) +
  
  #Reversing the direction of the legend
  guides(fill = guide_colorbar(reverse = TRUE)) 

#ggsave(one, file = "C:/Users/Ryan/Desktop/c123.png")
```


## Kent county map data for Yes Vote (all locales included)
```{r}
spatial[c(23, 12, 13), "NAME"] <- c("GR CT", "Lowell City", "Lowell CT")
  
ggplot() +
  geom_sf(data = kent, fill = "lightgrey", alpha = 0.5) +
  geom_sf(data = spatial, fill = "lightgrey", alpha = 0.5) +
  geom_sf(data = merged_data, aes(fill = percent_yes)) +
  
  #Adding labels to each county subdivision 
  geom_sf_text(data = spatial, aes(label = NAME), color = "black", size = 2.2) +
  
  scale_fill_distiller(palette = "Spectral", trans = "reverse") +
  
  labs(title = "Kent County Voting Data for KDL Millage Renewel",
       fill = "Percent of Yes Votes",
       caption = "Source: Kent Election Reporting") +
  
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), #hjust = 0.5 centers the title
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 8, colour = "black"),
        plot.caption = element_text(hjust = 1.5, vjust = 5)) +
  
  #Reversing the direction of the legend
  guides(fill = guide_colorbar(reverse = TRUE)) 

#ggsave(one, file = "C:/Users/Ryan/Desktop/c18723.png")
```



```{r eval=FALSE, include=FALSE}
#problem with state of michigan counties on the lakeshore

hmssdf <- filter(USA, STATEFP %in% c(26))

ggplot(hmssdf, fill = "red") +
  geom_sf()
```


```{r eval=FALSE, include=FALSE}
# Saving my merged df into a shape file (can't be excel, df is already saved as simple features (sf) df)

st_write(spatial, "C:/Users/Ryan/Desktop/kentelectiondate.shp")
```


# Alternate method of using the usmap library 
```{r eval=FALSE, include=FALSE}
library(usmap)

plot_usmap(regions = "counties") + 
  labs(title = "US Counties",
       subtitle = "This is a blank map of the counties of the United States.") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue"))

plot_usmap(data = statepop, values = "pop_2015", color = "red") + 
  scale_fill_continuous(name = "Population (2015)", label = scales::comma) + 
  theme(legend.position = "right")

usmap::plot_usmap("counties", 
                  include = c(.east_north_central))

#link with more info: https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html
```